Multi-level generalizations of collective spin dynamics and squeezing

### 2019-09-26 collective_sun/sun_notes.tex
added explicit derivation of SOC hamiltonian in drive operator basis

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{JM} \equiv \sqrt{\f{2J+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,JM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $J\in\set{0,1,\cdots,n-1}$ and $M\in\set{-J,-J+1,J}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of an $n$-level
spin; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,JM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{JM}|T_{J'M'}} = \delta_{JJ'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f12 u \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f12 u \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \v\T\equiv\sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{J,0} \equiv \sqrt{2}~ T_{J,0},
  &&
  D_{J\Delta,+} \equiv -\p{T_{J\Delta} + T_{J\Delta}^\dag},
  &&
  D_{J\Delta,-} \equiv i\p{T_{J\Delta} - T_{J\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for $\Delta\in\set{1,2,\cdots,J}$.  Similarly to the GGM operators in
\eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive operators
are equal to the Pauli operators in the case of $n=2$, and satisfy the
orthonormality condition
\begin{align}
  \obk{D_a|D_b} = 2\delta_{ab},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  These
operators have the convenient feature that $D_{1,1,\pm}$ is (up to a
gauge transformation) proportional to the Hamiltonian induced on an
$n$-level spin by a classical driving laser.  In terms of the drive
operators,
\begin{align}
  H_{\t{int}} = -\f14 u \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

We note that Weyl operators\cite{bertlmann2008bloch} and self-adjoint
linear combinations thereof\cite{asadian2016heisenbergweyl} also
provide an interesting basis of operators, generalizing the notion of
position and momentum to the case of a finite-dimensional Hilbert
space.  As our present work is motivated by multi-level systems that
are realized with nuclear spin (angular momentum) degrees of freedom,
however, the transition and drive operators are more natural to
consider than the Weyl operators, so we will not discuss the latter
any further.

@@

An external driving laser addressing a spin-$1/2$ particle on a
lattice will induce the Hamiltonian
\begin{align}
  \left. H_{\t{drive}}^{(\phi)} \right|_{n=2}
  = \f12 \Omega \sum_j e^{i\phi_j} c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  \label{eq:H_drive_2}
\end{align}
where $\Omega$ is a drive strength, $j$ indexes a lattice site, and
$\phi_j$ is a site-dependent phase.  The multi-level analogue of
\eqref{eq:H_drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)}
  = -\f12 \Omega \sum_j e^{i\phi_j} T_{1,1}^{(j)} + \t{h.c.}.
  \label{eq:H_drive}
\end{align}
We can make this drive spatially homogenous via the gauge
transformation $c_{j\mu} \to e^{i\mu\phi_j} c_{j\mu}$, after which the
drive becomes
\begin{align}
  H_{\t{drive}}
  = -\f12 \Omega \sum_j T_{1,1}^{(j)} + \t{h.c.}
  = \f12 \Omega \sum_j D_{1,1,+}
  = \Omega \D_{1,1,+},
\end{align}
where $\D_a\equiv\sum_jD_a^{(j)}/2$ is a collective drive operator,
with e.g.~$\D_{1,1,+}=S_\x$ in the case of $n=2$.  The SU($n$)
symmetry of collective interactions means that the interaction
Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_drive} remains unchanged
under this gauge transformation, but the single-particle Hamiltonian
in \eqref{eq:H_lat} transformed as
\begin{align}
  H_{\t{lat}}
  \to H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} c_{q\mu}^\dag c_{q\mu}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)},
\end{align}
where $\phi$ is the relative phase between adjacent lattice sites.  In
terms of the drive operators, we can expand
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -\sum_{q,J}\p{-1}^J f_J\p{qa} B_J\p{\phi} D_{J,0}^{(q)},
  &&
  f_J\p{\theta} \equiv
  \begin{cases}
    \cos\theta & J\,\t{even} \\
    \sin\theta & J\,\t{odd}
  \end{cases},
\end{align}
where the effective ``fields'' $B_J\p{\phi}$ are
\begin{align}
  B_J\p{\phi}
  = \f12 \sum_\mu \obk{D_{J,0}|s_{\mu\mu}} f_J\p{\mu\phi}
  = \sqrt{\f{J+1/2}{n}} \sum_\mu \bk{I\mu;I\mu,J,0} f_J\p{\mu\phi}.
\end{align}

@@

>>>>

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{LM} \equiv \sqrt{\f{2L+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,LM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $L\in\set{0,1,\cdots,n-1}$ and $M\in\set{-L,-L+1,L}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of an $n$-level
spin; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,LM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{LM}|T_{L'M'}} = \delta_{LL'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f12 u \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f12 u \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \v\T\equiv\sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{L,0} \equiv \sqrt{2}~ T_{L,0},
  &&
  D_{LM} \stackrel{M>0}{\equiv} -\p{T_{L\Delta} + T_{L\Delta}^\dag},
  &&
  D_{LM} \stackrel{M<0}{\equiv} i\p{T_{L\Delta} - T_{L\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for integer $M$ with $\abs{M}\le L$.  Similarly to the GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive
operators are equal to the Pauli operators in the case of $n=2$, and
satisfy the orthonormality condition
\begin{align}
  \obk{D_{LM}|D_{L'M'}} = 2\delta_{LL'}\delta_{MM'},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  These
operators have the convenient feature that $D_{1,1}$ is (up to a gauge
transformation) proportional to the Hamiltonian induced on an
$n$-level spin by a classical driving laser.  In terms of the drive
operators,
\begin{align}
  H_{\t{int}} = -\f14 u \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

We note that Weyl operators\cite{bertlmann2008bloch} and self-adjoint
linear combinations thereof\cite{asadian2016heisenbergweyl} also
provide an interesting basis of operators, generalizing continuous
phase-space displacement operators to the case of a finite-dimensional
Hilbert space.  As our present work is motivated by multi-level
systems that are realized with nuclear spin (angular momentum) degrees
of freedom, however, the transition and drive operators are more
natural to consider than the Weyl operators, so we will not discuss
the latter any further.

@@

An external driving laser addressing a spin-$1/2$ particle on a
lattice will induce the Hamiltonian
\begin{align}
  \left. H_{\t{drive}}^{(\phi)} \right|_{n=2}
  = \f12 \Omega \sum_j e^{i\phi_j} c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  \label{eq:H_drive_2}
\end{align}
where $\Omega$ is a drive strength, $j$ indexes a lattice site, and
$\phi_j$ is a site-dependent phase.  The multi-level analogue of
\eqref{eq:H_drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)}
  = -\f12 \Omega \sum_j e^{i\phi_j} T_{1,1}^{(j)} + \t{h.c.}.
  \label{eq:H_drive}
\end{align}
We can make this drive spatially homogenous via the gauge
transformation $c_{j\mu} \to e^{i\mu\phi_j} c_{j\mu}$, after which the
drive becomes
\begin{align}
  H_{\t{drive}}
  = -\f12 \Omega \sum_j T_{1,1}^{(j)} + \t{h.c.}
  = \f12 \Omega \sum_j D_{1,1}
  = \Omega \D_{1,1},
\end{align}
where $\D_{LM}\equiv\sum_jD_{LM}^{(j)}/2$ is a collective drive
operator, with e.g.~$\D_{1,1}=S_\x$ in the case of $n=2$.  The SU($n$)
symmetry of collective interactions implies that the interaction
Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_drive} is invariant under
this gauge transformation, but the single-particle Hamiltonian in
\eqref{eq:H_lat} transforms as
\begin{align}
  H_{\t{lat}}
  \to H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} c_{q\mu}^\dag c_{q\mu},
  \label{eq:H_lat_SOC}
\end{align}
where $\phi$ is the relative phase of the driving laser between
adjacent lattice sites.  In the basis of drive operators, this
Hamiltonian takes the form (see Appendix \ref{sec:lat_drive})
\begin{align}
  H_{\t{lat}}^{(\phi)} = -J \sum_{q,L} B_L\p{q,\phi} D_{L,0}^{(q)},
  \label{eq:H_lat_drive}
\end{align}
with effective driving field strengths
\begin{align}
  B_L\p{q,\phi} = \p{-1}^L f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu;I\mu,L,0} f_L\p{\mu\phi},
  &&
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spin-orbit coupling with drive operators}
\label{sec:lat_drive}

Here we take the single-particle spin-orbit coupled Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC}, and derive its
expression in \eqref{eq:H_lat_drive} in terms of the drive operators
$D_{LM}$.  In the basis of ordinary spin operators $s_{\mu\mu}$, this
Hamiltonian takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  \omega_{\phi,+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  \omega_{\phi,-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can simplify
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \omega_{\phi,+} - \sin\p{qa} \omega_{\phi,-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I \omega_{\phi,\pm} = \omega_{\phi,\pm}$ in the basis of the drive
operators, finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  - \sin\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $\omega_{\phi,-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0}
  - \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{\omega_{\phi,-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L B_L\p{q,\phi} D_{L,0},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field strengths
\begin{align}
  B_L\p{q,\phi}
  &\equiv \p{-1}^L f_L\p{qa} \f12
  \sum_\mu f_L\p{\mu\phi} \obk{s_{\mu\mu}|D_{L,0}} \\
  &= \p{-1}^L f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu;I\mu,L,0} f_L\p{\mu\phi},
\end{align}
where
\begin{align}
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = -J \sum_{q,L} B_L\p{q,\phi} D_{L,0}^{(q)}.
\end{align}
